---
title: "Extract and aggregate gridded datasets with R"
output: 
  html_notebook: 
    toc: yes
---

This [R Markdown](http://rmarkdown.rstudio.com) Notebook has two distinct goals:

* show how climatic data stored as ncdf format can easily be extracted in R;
* illustrate how gridded data can be aggregated at a given administrative level.

# Datasets

For this purpose, the gridded daily weather data for 1950-2015 available from the [E-OBS  repository](http://www.ecad.eu/download/ensembles/download.php) (v13.1 at 25 km x 25km horizontal resolution) is used (Haylock et al., 2008)[^refeobs].

[^refeobs]: Haylock, M.R., N. Hofstra, A.M.G. Klein Tank, E.J. Klok, P.D. Jones, M. New. 2008: A European daily high-resolution gridded dataset of surface temperature and precipitation. J. Geophys. Res (Atmospheres), 113.

The code chunk below reads the daily maximum temperature (**tx**) records from E-OBS and creates a RasterBrick i.e. a multi-layer raster object:
```{r eobs_read, echo=TRUE, message=FALSE, warning=FALSE}
require(raster)
require(magrittr)

eobs.path <- "~/Desktop/EOBS/%s_0.25deg_reg_v13.1.nc"
tmax.nc <- sprintf(eobs.path, 'tx') %>%
        brick()

tmax.nc
```

Plotting the first record (1950-01-01):
```{r eobs_plot, echo=TRUE, fig.height=4, fig.width=5, message=FALSE, warning=FALSE}
plot(tmax.nc[[1]])
```

The analysis is performed for France at the departement level (admin 2). The shapefile (shp) comes from the [Global Administrative Areas | Boundaries without limits](http://www.gadm.org/).

```{r shp, echo=TRUE, message=FALSE, warning=FALSE}
ISO <- 'FRA'
shp <- getData('GADM', country = ISO, level = 2)

shp
```

# Grid creation

The next step is the creation of the grid for E-OBS. Therefore, a mask with non NaNs is created and extracted for locations within the shp for France. Then, coordinates are extracted from the mask and saved as *SpatialPoints*.

```{r mask, echo=TRUE, fig.height=4, fig.width=4, message=FALSE, warning=FALSE}
require(dplyr)

mask.EOBS <- !is.na(tmax.nc[[1]]) %>%
        crop(shp) %>%
        mask(shp)

grid.EOBS <- coordinates(mask.EOBS) %>%
    as.data.frame() %>%
    mutate(keep = getValues(mask.EOBS)) %>%
    filter(keep) %>%
    SpatialPoints()
projection(grid.EOBS) <- projection(shp)

plot(grid.EOBS, col = 'red')
#plot(shp, add = T)
```

A lookup table is then created in which the admin levels 1 and 2 of each grid point is stored:
```{r lookup, echo=TRUE, message=FALSE, warning=FALSE}
lookup <- over(grid.EOBS, shp) %>%
    tbl_df()
lookup
```

# Time Series Extraction

The extraction of grid time series from a RasterBrick is fairly straighforward in R, by subsetting the RasterBrick with a SpatialPoints object (in this case the grid definition). This operation can be quite time consuming. The resulting data structure is a matrix, which is then transposed and converted to a data.frame (long format).

```{r ts, echo=TRUE, message=FALSE, warning=FALSE}
require(tidyr)
require(stringr)
require(lubridate)

tmax.ts <- tmax.nc[grid.EOBS] %>%
        t %>%
        as.data.frame() %>%
        mutate(date = row.names(.) %>% str_sub(start = 2) %>% ymd) %>%
        tbl_df %>%
        gather(key = id, value = tmax, -date) %>%
        mutate(id = id %>% str_sub(start = 2) %>% as.integer())

tmax.ts
```

# Use Case

This section contains a brief analysis of the number of critical days ($T_\mathit{max} > 34^\circ C$) per departement. First, the departement is added to each record (*lookup.adm2*).

```{r ts2, echo=TRUE, message=FALSE, warning=FALSE}
lookup.adm2 <- dplyr::select(lookup, NAME_2) %>%
    rename(departement = NAME_2) %>% 
    mutate(id = row.names(.) %>% as.integer())

tmax.ts.lookup <- tmax.ts %>%
    mutate(year = year(date),
           month = month(date)) %>% 
    left_join(lookup.adm2)
```

```{r crit, echo=TRUE, message=FALSE, warning=FALSE}
threshold <- 34.  # degC

crit.days <- tmax.ts.lookup %>%
    group_by(year, id, departement) %>%
    summarise(N = sum(tmax > threshold)) %>%
    ungroup() %>%
    group_by(year, departement) %>%
    summarise(N.avg = mean(N))
```

The map below shows the average number of critical days per departement:
```{r plot_map, echo=TRUE, fig.height=4, fig.width=5, message=FALSE, warning=FALSE}
require(ggplot2)
require(maptools)

shp.ggplot <- shp %>%
    fortify(region = 'NAME_2') %>%
    rename(departement = id)

dat.plot <- crit.days %>%
    group_by(departement) %>%
    summarise(N.avg = mean(N.avg)) %>%
    right_join(shp.ggplot)

ggplot(dat.plot) + 
    aes(long, lat, group = group, fill = N.avg) + 
    geom_polygon() +
    geom_path(color = "white") +
    scale_fill_gradient(low = 'grey', high = 'darkred') +
    theme_void()
```

The histogram below shows the distribution of critical days:
```{r plot_hist, echo=TRUE, message=FALSE, warning=FALSE}
ggplot(crit.days,  aes(x = year, y = N.avg, fill = departement)) +
    geom_bar(stat ="identity") +
    theme_bw() +
    theme(legend.position = "none") +
    xlab('') +
    ylab('# critical days')
```